Mass surveillance in cyberspace and the lost art of keeping a secret\thanks{The ideas discussed here first appeared as a DCSS project conference talk and were subsequently fleshed out in paper due to serendipitous encounters fostered by twitter.}

### 2015-07-08 edemocracy.tex
Tidying up folder, added Makefile, some changes
TC changes: tweaks, some text, refs and images

\documentclass{llncs}
\usepackage{graphicx,soul,color}

@@

\title{Mass surveillance in cyberspace and the lost art of keeping a secret\thanks{The ideas discussed here first appeared as a DCSS project conference talk and were subsequently fleshed out in the form of a paper due to serendipitous encounters that were lovingly fostered by the social medium of Twitter.}}
\subtitle{Policy lessons for nation-states after the Snowden leaks}

\author{Theo Tryfonas\inst{1} \and Michael Carter\inst{2} \and Tom Crick\inst{3} \and Panagiotis Andriotis\inst{1}}
\institute{Crypto Group, University of Bristol, UK\\
\email {t.tryfonas@bristol.ac.uk, p.andriotis@bristol.ac.uk}
\and
Surveillance Studies Centre, Queen's University, Canada\\
\email{michael.carter@queensu.ca} 
\and
Dept. of Computing, Cardiff Metropolitan University, UK\\
\email{tcrick@cardiffmet.ac.uk}}

@@

\begin{abstract}
Global security concerns, acts of terrorism and organised crime activity have motivated nation states to delve into implementing measures of mass surveillance in cyberspace, the breadth of which was partly revealed by the whistleblower Edward Snowden. But are modern nation states fighting a battle in the wrong space? Is mass surveillance of cyberspace effective and are the conventional metaphors of technology control appropriate for it? Can algorithms detect, classify and decide effectively on what constitutes suspicious activity? We argue that as cyberspace is a construct that has only recently been viewed strategically, let alone indoctrinated (the UK’s cyber-security strategy is only 4 years old), the societal impact of such bulk measures is yet much unclear – as are the assumptions about the fitness of state organisations that are charged with their oversight and the potential for unintended consequences. Recent experiences highlight the role of multiple forms of intelligence inputs, especially human- and community-based, and the need for application of such intrusive measures in a targeted manner. We believe that intrusive measures, where necessary, must be used decoupled from the seductive promises of advanced technology and ought to go hand-in-hand with means that strengthen the affected communities to identify, report and battle extremism and organised crime, in ways that safeguard the fundamental principles of our contemporary democratic Western states.
\keywords{Surveillance, cyberspace, public trust}
\end{abstract}

\section{Introduction}
\label{sec:Introduction}
In the fall of 2014, UN Special Rapporteur Ben Emmerson submitted his report on practices of mass surveillance by state actors and the threat that this approach to intelligence gathering poses to universal civil and political rights~\cite{Emerson}. Emmerson called for open and transparent discussion between government and citizens to inform and determine an appropriate balance between public security and personal privacy. The Special Rapporteur pointed out that what is technologically possible is not necessarily desirable or responsible. This is an argument that surveillance scholars such as Kirstie Ball have been making for several years now~\cite{Ball}. However, traction for this debate was limited until June 2013 when files leaked by NSA whistleblower Edward Snowden were published in the Guardian by journalist Glenn Greenwald.

Two years after the initial release of Snowden files, surveillance legislation remains highly contested in Canada, the US and the UK. Perhaps most notably is the sunsetting of section 215 of The Patriot Act and subsequent passing of The Freedom Act in the United States in early June 2015~\cite{Patriot}. Days later the Senate of Canada passed controversial anti-terrorism Bill C-51, which received sustained public opposition from big business, journalists, law professors, activists and the privacy commissioner~\cite{C-51}. A week prior to these developments the latest rendition of the so called 'snooper's charter' in the UK was announced in the Queen's speech. Former deputy Prime Minister Nick Clegg publicly opposed the legislation, currently known as The Investigatory Powers Bill, arguing it threatens the privacy rights of citizens. It's details are not fully fledged yet at the time of writing, but it is expected to be a contentious bill and Prime Minister David Cameron has at various points pointed towards banning the use of strong encryption -- albeit there would be fierce opposition to legislation against it.

These measures are indicative of state attempts to curb terrorism threats by enabling the development of surveillance capabilities that are of bulk collection nature, rather than targeted to specific individuals. Proponents of these argue that the proliferation of high technology, including anonymity, cryptography and secure communication tools, enables organised crime and terrorists, extremists etc. to communicate safely and go undetected. On the other hand privacy activists advocate the fundamental need for safe spaces to develop one's ideas, the human right to privacy and an individual's need to protect themselves from abusive regimes.

@@

\section{Background}
\label{sec:Background}
The debate on mass surveillance, which is comprised of several threads, has engaged a range of social groups including politicians, law makers, journalists, academics, tech firms, activists, artists and the general public. The term mass surveillance is used to distinguish the bulk collection of data from targeted surveillance, which typically involves a 'person of interest'. Central to this aspect of the debate is the legal warrant, which is traditionally issued upon satisfaction of a certain level of suspicion. In the case of Canada, for example, Bill C-13, which was passed in the fall of 2014, significantly lowered the level of suspicion required to justify the collection of personal data. Bill C-13 also addressed the distinction between data and meta data, which is a hotly debated topic in surveillance legislation. In particular, Bill C-13 reduces the limitations on collecting meta data under the guise that it is not intrusive. Advocates of expanded surveillance powers for the state have attempted to mollify concerns by arguing that meta data does not threaten the political or civil rights of citizens because it is data about communication and not the content of communication. This argument has been routinely problematized by opponents who point out that metadata can reveal religious beliefs, political leanings and intimate relationships. Moreover, meta data is used by state actors to kill people, as was famously announced by former NSA and CIA director Michael Hayden~\cite{Hayden}.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1}
\caption{Yahoo! webcam traffic monitoring report snapshots from Snowden's cache.}
\label{fig:Yahoo}
\end{center}
\end{figure}

The Snowden leaks revealed a wide portfolio of projects and initiatives both from the NSA in the US, CSIS in Canada and GCHQ in the UK. These range from specific data collection projects such as Optic Nerve, aimed at Yahoo! webcam traffic (Figure \ref{fig:Yahoo}), to influencing the development of cryptographic standards to contain vulnerabilities, so they can be penetrated easier~\cite{ECDH}. In this varied context the Anderson report~\cite{Anderson} that was released recently as a comprehensive review of the UK's capabilities and practice prior to revamping the existing legislation, emphasised a number of issues, amongst the most important - and contested - of which, was the suggestion for judicial rather than ministerial oversight.

@@

Another frequently used analogy is the case of the closed-circuit television (CCTV) surveillance systems. This is a familiar, and very tangible, system which in the UK at least enjoys large amounts of public tolerance and even approval~\cite{Ditton}, even at the face of lack of real evidence of its effectiveness~\cite{Woodhouse}. We will get back to this a bit later, discussing the experience of insitutionalisation of CCTV as a means of surveillance, particularly in the UK where it is widely deployed across the country.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.3]{fig2}
\caption{Part of Jeremy Bentham's designs for a Panopticon prison.}
\label{fig:Panopticon}
\end{center}
\end{figure}

Whether the Panotpicon metaphor matches the underlying security requirements is a significant question. This is because a metaphor is a conceptual construct able to shape action, as demonstrated by several scholars, including e.g. Tsoukas~\cite{Tsoukas}. Very soon after the attacks of 9/11, Lackoff argued that inappropriately framing the reaction as a 'war on terror' would produce unintended consequences~\cite{Lackoff}. Other research shows how secure systems implementation is shaped by the dominant security metaphors in use within organisations~\cite{Tryfonas}.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig3}
\caption{Failures of algorithmic determination after the Boston bombings.}
\label{fig:Algo}
\end{center}
\end{figure}

Another implicit assumption to legitimise this view is that this activity is organised under a framework of strong oversight. Particularly for the Anglo-Saxon world and especially for the UK, in the light of the strong leadership of prominent politicians such as Winston Churchill, Margaret Thatcher and Tony Blair, this assurance is almost taken for granted. However, history suggests that in the absence of oversight, socio-political circumstances may provide opportunity for exploitation of such structures. The experience of the rise of the Nazi party in the Weimar Republic is in line with this observation. Finally, we often also forget that the Internet is in reality a young technological development that it is yet much unexplored in terms of national security and military doctrine and use. The UK Cyber Security Strategy for example is less than 5 years old~\cite{Cyber}.

@@

Regardless of the heated debate about the effectiveness of CCTV against crime~\cite{Ditton}\cite{Woodhouse}, its proliferation and acceptance in some societies is notable. This is certainly the case in the UK, where CCTV rose to public prominence largely in the 80's. Part of its take up may be due to the campaign against football hooliganism as well as ground safety fears after the tragedy of Hillsborough. Some researchers also suggest that they became instruments of enforcing an image of a tough stance against crime by political parties, especially New Labour~\cite{McHill}.

@@

Translating this into the cyber domain, there is a need for related technology to be viewed as an enabler, as opposed to being demonised and fundamentally mistrusted. Greater transparency of Internet surveillance programmes and sufficient oversight structures would assure the public of the role of technology. Further education and public understanding of surveillance would help. This is not incompatible with the secrecy that security services claim must surround their operations; in fact they ought to assume that adversaries are suspicious of their practices. The Panotpticon metaphor needs to be revisited as until now it has blurred the realisation of the need to intervene earlier in the radicalisation lifecycle to debunk their propaganda messages and the appeal of radicalism to young and vulnerable members of society. There is a need for more human-centric intelligence, open source and targeted operations, as opposed to passive monitoring and algorithmic determination implied by the current paradigm. We believe that if the purpose is clear and the community is able to see the value to them, they will even tolerate the trade off of personal data in return for confronting effectively this threat, as suggested even by the commercial experiences discussed earlier.

@@

\bibitem{Patriot}
Kelly, E.:
Senate approves USA Freedom Act, Usa Today,  June 2, 2015

@@

\bibitem{Hayden}
Ferran, L.:
Ex-NSA Chief: 'We Kill People Based on Metadata', abcNEWS, May 12, 2014

@@

\bibitem{Cyber}
UK Government:
The UK Cyber Security Strategy: Protecting and promoting the UK in a digital world, 2011

\bibitem{GoogleAds}
Google Inc.:
http://www.google.com/ads/

>>>>

\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage[british]{babel}
\usepackage{url}
\usepackage{graphicx}
\usepackage{soul}
\usepackage{color}
\usepackage[pdftex,colorlinks=true]{hyperref}

@@

\title{Mass surveillance in cyberspace and the lost art of keeping a secret\thanks{The ideas presented in this paper first appeared as a talk at the 2015 International Conference on Digital Citizenship and Surveillance Society in Cardiff in June 2015 and subsequently refined and developed into this paper via serendipitous encounters that were enabled by the social medium of Twitter.}}
\subtitle{Policy lessons for nation-states after the Snowden leaks}

\author{Theo Tryfonas\inst{1} \and Michael Carter\inst{2} \and Tom Crick\inst{3} \and Panagiotis Andriotis\inst{1}}
\institute{Crypto Group, University of Bristol, UK\\
\email {\{t.tryfonas,p.andriotis\}@bristol.ac.uk}
\and
Surveillance Studies Centre, Queen's University, Canada\\
\email{michael.carter@queensu.ca} 
\and
Dept. of Computing, Cardiff Metropolitan University, UK\\
\email{tcrick@cardiffmet.ac.uk}}

@@

\begin{abstract}
Global security concerns, acts of terrorism and organised crime activity have motivated nation states to delve into implementing measures of mass surveillance in cyberspace, the breadth of which was partly revealed by the whistleblower Edward Snowden. But are modern nation states fighting a battle in the wrong space? Is mass surveillance of cyberspace effective and are the conventional metaphors of technology control appropriate for it? Can algorithms detect, classify and decide effectively on what constitutes suspicious activity? We argue that as cyberspace is a construct that has only recently been viewed strategically, let alone indoctrinated (the UK’s cyber-security strategy is only four years old), the societal impact of such bulk measures is yet much unclear -- as are the assumptions about the fitness of state organisations that are charged with their oversight and the potential for unintended consequences. Recent experiences highlight the role of multiple forms of intelligence inputs, especially human- and community-based, and the need for application of such intrusive measures in a targeted manner. We believe that intrusive measures, where necessary, must be used decoupled from the seductive promises of advanced technology and ought to go hand-in-hand with means that strengthen the affected communities to identify, report and battle extremism and organised crime, in ways that safeguard the fundamental principles of our contemporary democratic Western states.
\keywords{Surveillance, cyberspace, public trust}
\end{abstract}

\section{Introduction}
\label{sec:Introduction}
In the fall of 2014, UN Special Rapporteur Ben Emmerson submitted his report on practices of mass surveillance by state actors and the threat that this approach to intelligence gathering poses to universal civil and political rights~\cite{Emerson}. Emmerson called for open and transparent discussion between government and citizens to inform and determine an appropriate balance between public security and personal privacy. The Special Rapporteur pointed out that what is technologically possible is not necessarily desirable or responsible. This is an argument that surveillance scholars such as Kirstie Ball have been making for several years now~\cite{Ball}. However, traction for this debate was limited until June 2013 when files leaked by NSA whistleblower Edward Snowden were published in the Guardian by journalist Glenn Greenwald~\cite{Greenwald}.

Two years after the initial release of Snowden files, surveillance legislation remains highly contested in Canada, the US and the UK. Perhaps most notably is the sunsetting of section 215 of The Patriot Act and subsequent passing of the USA FREEDOM Act\footnote{\url{https://www.congress.gov/bill/113th-congress/house-bill/3361}} in the United States in early June 2015~\cite{Patriot}. Days later the Senate of Canada passed controversial anti-terrorism Bill C-51\footnote{\url{http://www.parl.gc.ca/HousePublications/Publication.aspx?DocId=6932136&Col=1&File=4}}, which would amend a number of other acts, as well as enact the Security of Canada Information Sharing Act, receiving sustained public opposition from big business, journalists, law professors, activists and the privacy commissioner~\cite{C-51}. A week prior to these developments the latest rendition of the so called `snooper's charter' in the UK was announced in the 2015 Queen's Speech. Former deputy Prime Minister Nick Clegg publicly opposed the legislation, currently known as the Investigatory Powers Bill\footnote{\url{https://wiki.openrightsgroup.org/wiki/Investigatory_Powers_Bill}}, arguing it threatens the privacy rights of citizens. Its details are not fully fledged yet at the time of writing, but it is expected to be a contentious bill and Prime Minister David Cameron has at various points pointed towards banning the use of strong encryption -- albeit there would be fierce opposition to legislation against it.

These measures are indicative of state attempts to curb terrorism threats by enabling the development of surveillance capabilities that are of bulk collection nature, rather than targeted to specific individuals. Proponents of these argue that the proliferation of high technology, including anonymity, cryptography and secure communication tools, enables organised crime and terrorists, extremists etc. to communicate safely and go undetected. On the other hand privacy activists advocate the fundamental need for safe spaces to develop one's ideas, the human right to privacy and an individual's need to protect themselves from abusive regimes. There is thus a significant research -- and national security -- focus on analysing and profiling digital and online behaviour~\cite{gosling:2009,oatley+crick:2014}.

Cyber security is thus emerging as one of the most challenging aspects of the information age for policymakers, technologists and scholars of international relations. It has implications for national security, the economy, human rights, civil liberties and international legal frameworks. Although politicians have been aware of the threats of cyber insecurity since the early years of Internet technology, anxiety about the difficulties in resolving or addressing them has increased rather than abated~\cite{obama:2009,bisinfosecbreaches:2014,keepingukcybersafe:2014}. In response, governments have begun to develop robust (and wide-reaching) national cyber security strategies~\cite{gwbush:2003,ukcss:2011,ukcss:2014} to outline the way in which they intend to address cyber insecurity. However, in many states where critical infrastructure such as utilities, financial systems and transport have been privatised, these policies are heavily reliant upon what is referred to as the `public-private partnership' as a key mechanism through which to mitigate the threat. In the UK and US, the public-private partnership has repeatedly been referred to as the `cornerstone' or `hub' of cyber security strategy~\cite{carr+crick-csss2015}, which further raises questions regarding mass surveillance and data retention, particularly who has access to private citizens' data.

@@

\section{Background}
\label{sec:Background}
The debate on mass surveillance, which is comprised of several threads, has engaged a range of social groups including politicians, law makers, journalists, academics, tech firms, activists, artists and the general public. The term mass surveillance is used to distinguish the bulk collection of data from targeted surveillance, which typically involves a `person of interest'. Central to this aspect of the debate is the legal warrant, which is traditionally issued upon satisfaction of a certain level of suspicion. In the case of Canada, for example, Bill C-13, which was passed in the fall of 2014, significantly lowered the level of suspicion required to justify the collection of personal data. Bill C-13 also addressed the distinction between data and meta data, which is a hotly debated topic in surveillance legislation. In particular, Bill C-13 reduces the limitations on collecting meta data under the guise that it is not intrusive. Advocates of expanded surveillance powers for the state have attempted to mollify concerns by arguing that meta data does not threaten the political or civil rights of citizens because it is data about communication and not the content of communication. This argument has been routinely problematized by opponents who point out that metadata can reveal religious beliefs, political leanings and intimate relationships. Moreover, meta data is used by state actors to kill people, as was famously announced by former NSA and CIA director Michael Hayden~\cite{Hayden}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{images/fig1}
\caption{Yahoo! webcam traffic monitoring report snapshots from Snowden's cache.}
\label{fig:Yahoo}
\end{center}
\end{figure}

The Snowden leaks revealed a wide portfolio of projects and initiatives both from the NSA in the US, CSIS in Canada and GCHQ in the UK. These range from specific data collection projects such as Optic Nerve, aimed at Yahoo! webcam traffic (Figure~\ref{fig:Yahoo}), to influencing the development of cryptographic standards to contain vulnerabilities, so they can be penetrated easier~\cite{ECDH}. In this varied context the Anderson report~\cite{Anderson} that was released recently as a comprehensive review of the UK's capabilities and practice prior to revamping the existing legislation, emphasised a number of issues, amongst the most important - and contested - of which, was the suggestion for judicial rather than ministerial oversight.

@@

Another frequently used analogy is the case of the closed-circuit television (CCTV) surveillance systems. This is a familiar, and very tangible, system which in the UK at least enjoys large amounts of public tolerance and even approval~\cite{Ditton}, even at the face of lack of real evidence of its effectiveness~\cite{Woodhouse}. We will get back to this a bit later, discussing the experience of institutionalisation of CCTV as a means of surveillance, particularly in the UK where it is widely deployed across the country.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{images/fig2}
\caption{Part of Jeremy Bentham's designs for a Panopticon prison.}
\label{fig:Panopticon}
\end{center}
\end{figure}

Whether the Panopticon metaphor matches the underlying security requirements is a significant question. This is because a metaphor is a conceptual construct able to shape action, as demonstrated by several scholars, including e.g. Tsoukas~\cite{Tsoukas}. Very soon after the attacks of 9/11, Lackoff argued that inappropriately framing the reaction as a 'war on terror' would produce unintended consequences~\cite{Lackoff}. Other research shows how secure systems implementation is shaped by the dominant security metaphors in use within organisations~\cite{Tryfonas}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{images/fig3}
\caption{Failures of algorithmic determination after the Boston bombings.}
\label{fig:Algo}
\end{center}
\end{figure}

Another implicit assumption to legitimise this view is that this activity is organised under a framework of strong oversight. Particularly for the Anglo-Saxon world and especially for the UK, in the light of the strong leadership of prominent politicians such as Winston Churchill, Margaret Thatcher and Tony Blair, this assurance is almost taken for granted. However, history suggests that in the absence of oversight, socio-political circumstances may provide opportunity for exploitation of such structures. The experience of the rise of the Nazi party in the Weimar Republic is in line with this observation. Finally, we often also forget that the Internet is in reality a young technological development that it is yet much unexplored in terms of national security and military doctrine and use; for example, the UK's Cyber Security Strategy is less than five years old~\cite{ukcss:2011}.

@@

Regardless of the heated debate about the effectiveness of CCTV against crime~\cite{Ditton,Woodhouse}, its proliferation and acceptance in some societies is notable. This is certainly the case in the UK, where CCTV rose to public prominence largely in the 80's. Part of its take up may be due to the campaign against football hooliganism as well as ground safety fears after the tragedy of Hillsborough. Some researchers also suggest that they became instruments of enforcing an image of a tough stance against crime by political parties, especially New Labour~\cite{McHill}.

@@

Translating this into the cyber domain, there is a need for related technology to be viewed as an enabler, as opposed to being demonised and fundamentally mistrusted. Greater transparency of Internet surveillance programmes and sufficient oversight structures would assure the public of the role of technology. Further education and public understanding of surveillance would help. This is not incompatible with the secrecy that security services claim must surround their operations; in fact they ought to assume that adversaries are suspicious of their practices. The Panopticon metaphor needs to be revisited as until now it has blurred the realisation of the need to intervene earlier in the radicalisation lifecycle to debunk their propaganda messages and the appeal of radicalism to young and vulnerable members of society. There is a need for more human-centric intelligence, open source and targeted operations, as opposed to passive monitoring and algorithmic determination implied by the current paradigm. We believe that if the purpose is clear and the community is able to see the value to them, they will even tolerate the trade off of personal data in return for confronting effectively this threat, as suggested even by the commercial experiences discussed earlier.

@@

\bibitem{Greenwald}
Greenwald, G.:
No Place to Hide: Edward Snowden, the NSA and the Surveillance State, Hamish Hamilton, 2014

\bibitem{Patriot}
Kelly, E.:
Senate approves USA Freedom Act, {USA Today},  June 2, 2015

\bibitem{gosling:2009}
Gosling, S.:
\newblock {Snoop: What Your Stuff Says About You}.
\newblock {Profile Books} (2009)

\bibitem{oatley+crick:2014}
Oatley, G. and Crick, T.:
\newblock {Changing Faces: Identifying Complex Behavioural Profiles}.
\newblock In {\em {Human Aspects of Information Security, Privacy and Trust}},
  volume 8533 of {\em {Lecture Notes in Computer Science}}, pages 282--293.
  Springer, 2014.

\bibitem{obama:2009}
Barack Obama.
\newblock {Remarks by the President On Securing Our Nation's Cyber
  Infrastructure}.
\newblock
  \url{http://www.whitehouse.gov/the_press_office/Remarks-by-the-President-on-Securing-Our-Nations-Cyber-Infrastructure/},
  May 2009.

\bibitem{bisinfosecbreaches:2014}
{Department for Business, Innovation \& Skills}:
\newblock {2014 Information Security Breaches Survey}.
\newblock {UK Government}, 2014.

\bibitem{gwbush:2003}
George~W. Bush.
\newblock {The National Strategy to Secure Cyberspace}.
\newblock {The White House}, February 2003.

\bibitem{ukcss:2011}
{Cabinet Office}:
\newblock {Cyber Security Strategy}.
\newblock {UK Government}, November 2011.

\bibitem{ukcss:2014}
{Cabinet Office}.
\newblock {National Cyber Security Strategy 2014: progress and forward plans}.
\newblock {UK Government}, December 2014.

\bibitem{keepingukcybersafe:2014}
{Cabinet Office}.
\newblock {Keeping the UK safe in cyber space}.
\newblock {UK Government}, December 2014.

\bibitem{carr+crick-csss2015}
Carr, M. and Crick, T:
\newblock {The Problem of the P3: Public-Private Partnerships in National Cyber
  Security Strategies}.
\newblock In {\em {Proceedings of 1st International Conference on Cyber
  Security for Sustainable Society 2015}}.

@@

\bibitem{Hayden}
Ferran, L.:
Ex-NSA Chief: `We Kill People Based on Metadata', abcNEWS, May 12, 2014

@@

\bibitem{GoogleAds}
Google Inc.:
\url{http://www.google.com/ads/}
